Vestibular Contributions to Spatial Perspective Transformations: From Sensory Inference To Imagined Self-Motion

### 2017-01-27 draft-manuscript.Rmd
added Notes, model-fits and new content

Mental imagery is oftern portrayed as being the result of top-down reconstruction of activity in perceptual circuits in the absence of physical stimulation of the sensors, and is opposed to bottom-up perception. However, in the case of the vestibular system, it becomes clear that this view is inadequate (the same is true for other modalities): the sensory that the vestibular sensors provide is noisy and ambiguous (examples), and in order to deal with these facts, the brain must use several strategies (which are well-known from control theory). It is assumed that the brain uses (at least approximatively) statistically optimal filtering algorithms. Initia research cast the type of processing in the framework of control theory (optimal observer models); however @Selva:2012gsa showed that these models are equivalent to statistically optimal models (Kalman filters), in the case of linear dynamical systems.

@@

Head direction cells are thought to integrate signals of vestibular origin to maintain a signal of cumulative rotation (reviewed in @Taube:2007fk and are found in many brain areas, including the postsubiculum, retrosplenial cortex, thalamus, lateral mammillary nucleus, dorsal tegmental nucleus, striatum and entorhinal cortex [@Cullen:2014gx]. This may be necessary in order to mentally simulate being in another position, i.e. the brain apparently needs to simulate the n-step dynamics of its motion through space. Why is this? Because vestibular connections make a major contribution to spatial orientation; information about the position in space obtained through the vestibular sensory system is computed by integrating velocity over time.

>>>>

> Mental imagery is oftern portrayed as being the result of top-down reconstruction of activity in perceptual circuits in the absence of physical stimulation of the sensors, and is opposed to bottom-up perception. However, in the case of the vestibular system, it becomes clear that this view is inadequate (the same is true for other modalities): the sensory that the vestibular sensors provide is noisy and ambiguous (examples), and in order to deal with these facts, the brain must use several strategies (which are well-known from control theory). It is assumed that the brain uses (at least approximatively) statistically optimal filtering algorithms. Initia research cast the type of processing in the framework of control theory (optimal observer models); however @Selva:2012gsa showed that these models are equivalent to statistically optimal models (Kalman filters), in the case of linear dynamical systems.

@@

~Head direction cells are thought to integrate signals of vestibular origin to maintain a signal of cumulative rotation (reviewed in @Taube:2007fk and are found in many brain areas, including the postsubiculum, retrosplenial cortex, thalamus, lateral mammillary nucleus, dorsal tegmental nucleus, striatum and entorhinal cortex [@Cullen:2014gx]. This may be necessary in order to mentally simulate being in another position, i.e. the brain apparently needs to simulate the n-step dynamics of its motion through space. Why is this? Because vestibular connections make a major contribution to spatial orientation; information about the position in space obtained through the vestibular sensory system is computed by integrating velocity over time.~

### 2017-01-27 manuscript.Rmd
added Notes, model-fits and new content

---
# title             : "Modeling Cognitive Effects Vestibular Sensory Inference: From Sensory Inference To Imagined Self-Motion"

title             : "Vestibular Contributions to Spatial Perspective Transformations: From Sensory Inference To Imagined Self-Motion"

# title             : "A Computational Perspective On Spatial Perspective Transformations: From Sensory Inference To Imagined Self-motion"
# title             : "Probabilistic Computations Underlying Simulated Self-Rotations"
# title: Probabilistic Computations Underlying Vestibular Cognition
shorttitle        : "Vestibular Contributions to SPT"

@@

keywords: probabilistic graphical models, particle filter, embodied cognition, vestibular cognition, spatial perspective taking, mental imagery
wordcount: "X"

@@

figsintext        : yes
figurelist        : yes
tablelist         : yes
footnotelist      : no
number_sections   : no
lineno            : yes
linkcolor         : "blue"
# citeproc          : yes
# csl               : apa.csl

lang              : "english"
class             : "man"
output:
     # papaja::apa6_word:
     papaja::apa6_pdf:
        pandoc_args:
        fig_caption: true
---

```{r preliminary, cache=FALSE, message = FALSE, warning = FALSE}
library(papaja)
library(pander)
library(ggplot2)
library(dplyr)
library(tidyr)
library(mvtnorm)
# library(ggthemes)
library(cowplot)

@@

Adopting the spatial perspective of another person is generally considered an essential cognitive ability, and plays a vital role in our ability to determine what another person can see and in order to coordinate joint actions [@CreemRegehr:2013kx, @Pezzulo:2013iy]. Many studies have also pointed out that perspective taking is of fundamental importance for language comprehension and communication [@Beveridge:2013gx], and social cognition in general [@Deroualle:2014ho].

@@

Performing a simulation of body movements and estimating the sensory consquences amounts to simulating the behaviour of a dynamical system; this requires a generative model of the system. The aim of this paper is to specifically investigate the generative model that may be used for mental spatial transformations. The ability to run embodied simulations is often claimed to be grounded in the relevant sensorimotor circuits, and we claim that the vestibular system is ideally suited to investigate this connection, because the generative models used by the vestibular system for dynamic sensory inference have been extensively investigated using approaches from optimal control theory, and more recently, dynamic Baysian inference. The vestibular system faces a number of challenges in interpreting the sensory measurements provided by the semi-circular canals (SCC) and otolith organs. On the one hand, these signals  are very noisy, and ambiguous [@PaulRMacNeilage:2008bv]; the otoliths respond both to translations and tilt of the head relative to gravity, and in order to disambiguate between translation and tilt, the brain must perform sophisticated computations at the level of the brain stem. It has become clear that the nature of these computations is probabilistic, and that the brain makes use of strong prior beliefs in order to interpret these sensory signals. A further challenge faced by the vestibular system is that it needs to distinguish between sensory signals that are the result of active motion, and those resulting from external forces [@Cullen:2009dc, @Cullen:2014gx]. The ability to distinguishing re-afferent from ex-afferent sensory signals is an essential capability of all organisms that move actively [@Crapse:2008jc].

@@

Interference effects between higher-level cognitive processes, such as SPT and mental imagery, and sensory inference, are thought to occur as a result of shared circuitry [@Deroualle:2014ho, @Nigmatullina:2015gh]. @Ellis:2017fg recently claimed that these types of interference effects might arise as a result of cognitive processing exerting an influence on circuits performing sensory inference via shared higher-level priors. In this paper, we pursue this claim, and we attempt to the grounding of cognitive abilities in the computations performed by low-level sensory circuits in the service of dynamic real-time interactions with the world.

@@

@@

## Previous work
Whereas early models of vestibular sensory processing were inspired by control theory [@Borah:1988fd, @Merfeld:1999cg], several recent studies have described vestibular sensory processing as dynamic Bayesian inference [@Laurens:2006be, @PaulRMacNeilage:2008bv, @Karmali:2012cv]. @Selva:2012gsa
provide a recent review of the relationship between optimal observer models and models which employ Kalman filtering, and show that these are largely equivalent. Both @Karmali:2012cv and @Laurens:2006be described vestibular sensory inference in terms of particle filtering, which is a particular inference algorithm that can be used to perform inference in state-space models. [@Doucet:2000bh, @Doucet:2009us]. Using their model, @Laurens:2006be were able to simulate perceptual responses to centrifugation and off-vertical axis rotations. A well-known phenomenon of vestibular perception is velocity storage, which is defined as the slower decay of perceptual and oculomotor responses, relative to the SCC afferent responses [@Raphan:1979hn], __[Robinson 1977]__. This is generally seen as evidence for the existence of an internal model in the processing of vestibular signals. @Karmali:2012cv focussed on one-dimensional angular velocity during yaw rotations; their model was able to explain the phenomenon of velocity storage as a consequence of particle filtering, with the time constant of velocity storage depending on the variability of the sensory afferent signals.

## Generative model for sensory inference
We proceed by providing a brief introduction to the concept of state-space models, and particle filtering, which is a particular algorithm that can be used to perform inference in state-space models. While Kalman filters provide the optimal solution when the restrictions of linear Gaussian State-space models are met, they are not easily extensible to non-linear models, which is relevant when considering multi-dimensional motion. Particle filtering can be applied to arbitrary nonlinear models. State-space models can be seen as special cases of more general Bayesian hierarchical models [@Bishop:2006ui; @Murphy:2012ua]. Following the same approach as @Penny:2013iv, we first reframe the generative model used for sensory inference as a dynamic probabilistic graphical model
 <!-- from a statistical perspective, rather an engineering perspective -->
A state-space model is special case of a dynamic latent variable model which is suitable for modelling the time-varying behaviour of a system in terms of latent, unobservable processes and noisy observations. As an example, consider trying to estimate the velocity and position during a leftward head turn. The afferent SCC signals are considered to be proportional to the angular velocity of the head with a frequency band of approximately 0.1 – 10 Hz, which constitutes the range of natural head movements [@Grabherr:2008kk; @Carriot:2014fv]. The measurements are therefore observations of the true, unobservable angular velocity; however, they are contaminted by Gaussian noise, which is determined by characteristics of the sensory organs. Figure \@ref(fig:head-turn-data) shows a sequence of noisy measurements during a two-second turn of the head. The acceleration generating the movement is sinusoidal, given for each time step $t$ by $Asin(2\pi t/T)$, where $A$ is the amplitude of the movement, $T = 1/f$ is the duration of the movement, and $f$ is the frequency. This acceleration results in a peak velocity $\omega_{peak} = AT/\pi$ and a final angular position $\theta_{final} = AT^2/2\pi$.  Due to the inherent noisiness of the measurements, and the dynamic nature of the stimulus, the measurements need to be interpreted according to the context in which they occur.

@@

<!-- \autoref{fig:graphical-model-inference} -->
\@ref(fig:graphical-model-inference) shows a representation of a state-space model as a graphical model, unrolled over multiple time steps. Starting at time 0, the state evolves according to the process model, with the measurements at each time step being generated according to the measurement model. The 2-dimensiontal latent state variable is unshaded, to indicate that it cannot be observed, wheres the observed sensory measurements are shaded. Whereas previous models were discussed in terms of the particular filtering algorithm used, the graphical notation highlights the fact that a state-space model can be viewed as a dynamic latent factor model [@Blei:2014cp]. This opens up the possibility of extending the graphical model to include higher-level variables. 
Given a generative model, we can apply various sampling algorithms, either in order to run a Monte Carlo simulation by sampling from the prior without incorporating any observed variables, or to perform inference, by conditioning on observed variables. After a brief discussion particle filtering, an inference algorithm often used to infer the values of the latent state variables, we will introduce our extension to the model that will allow a prediction of the consequences of active movements.

@@

$$ \omega_{t} = \omega_{t-1} + \alpha_{t-1} \Delta t $$

@@

@@

@@

#TODO: model implementation

@@

@@

- ability to select type of motion
- parameters of tht motion

@@

# Various sources of interference
Here, we identify various sources of intereference (2-way) between thought (higher-level cognition) and lower-level sensory processing. This interference can be either due to low-level influences in sensory inference, due to priors on higher-level parameters, choice of process model (false expectations of certain type of motion). This is shown by model simulations.

@@

## Summary
Interference may the result of either inappropriate priors, leading to a mis-specified filtering algorithm, or biased decision-making. We should be able to investigate which by applying cognitive process models that explain response times and choices.

@@

@@

- Nigmatullina
- Wertheim

@@

### Head direction cells
Why dynamic simulation? Maybe in order to provide vestibular in put to head direction cells [@Taube:2007fk]

>>>>

---
# title: "Modeling Cognitive Effects Vestibular Sensory Inference: From Sensory Inference To Imagined Self-Motion"

title: "Vestibular Contributions to Spatial Perspective Transformations: From Sensory Inference To Imagined Self-Motion"

# title: "A Computational Perspective On Spatial Perspective Transformations: From Sensory Inference To Imagined Self-motion"
# title: "Probabilistic Computations Underlying Simulated Self-Rotations"
# title: Probabilistic Computations Underlying Vestibular Cognition
shorttitle        : "Vestibular Contributions to SPT"

@@

keywords: probabilistic graphical models, particle filter, embodied cognition, vestibular cognition, spatial perspective taking, mental imagery, mental simulation

wordcount: "X"

@@

figsintext        : yes
figurelist        : yes
tablelist         : yes
footnotelist      : no
number_sections   : no
lineno            : yes
linkcolor         : "blue"
fig_caption       : true

# citeproc          : yes
# csl               : apa.csl

lang              : "english"
class             : "man"
output:
     # papaja::apa6_word:
     papaja::apa6_pdf
        # pandoc_args:
        # - --filter
        #   pandoc-fignos

---

```{r preliminary, cache=FALSE, message = FALSE, warning = FALSE}
library(papaja)
# library(pander)
library(ggplot2)
library(dplyr)
library(tidyr)
library(mvtnorm)
# library(ggthemes)
library(cowplot)

@@

Adopting the spatial perspective of another person is an essential cognitive ability, and plays a vital role in our ability to determine possible actions of another person, and in order to coordinate joint actions [@CreemRegehr:2013kx, @Pezzulo:2013iy]. Perspective taking is also considered to be of fundamental importance for language comprehension and communication [@Beveridge:2013gx], and social cognition in general [@Deroualle:2014ho].

@@

Performing a simulation of body movements and estimating the sensory consquences amounts to simulating the behaviour of a dynamical system using a forward model; this requires a generative model of the system.The aim of this paper is to specifically investigate the generative model that may be used for mental spatial transformations. The ability to run embodied simulations is often claimed to be grounded in the relevant sensorimotor circuits, and we claim that the vestibular system is ideally suited to investigate this connection, because the generative models used by the vestibular system for dynamic sensory inference have been extensively investigated using approaches from optimal control theory, and more recently, dynamic Baysian inference. The vestibular system faces a number of challenges in interpreting the sensory measurements provided by the semi-circular canals (SCC) and otolith organs. On the one hand, these signals  are very noisy, and ambiguous [@PaulRMacNeilage:2008bv]; the otoliths respond both to translations and tilt of the head relative to gravity, and in order to disambiguate between translation and tilt, the brain must perform sophisticated computations at the level of the brain stem. It has become clear that the nature of these computations is probabilistic, and that the brain makes use of strong prior beliefs in order to interpret these sensory signals. A further challenge faced by the vestibular system is that it needs to distinguish between sensory signals that are the result of active motion, and those resulting from external forces [@Cullen:2009dc, @Cullen:2014gx]. The ability to distinguishing re-afferent from ex-afferent sensory signals is an essential capability of all organisms that move actively [@Crapse:2008jc].

@@

Interference effects between higher-level cognitive processes, such as SPT and mental imagery, and sensory inference, are thought to occur as a result of shared circuitry [@Deroualle:2014ho, @Nigmatullina:2015gh]. @Ellis:2017fg recently claimed that these types of interference effects might arise as a result of cognitive processing exerting an influence on circuits performing sensory inference via shared higher-level priors. In this paper, we pursue this claim, and we attempt to the grounding of cognitive abilities in the computations performed by low-level sensory circuits in the service of dynamic real-time interactions with the world. A problem that needs to addressed is whether models that are applicable to lower-level sensory inference are scalable, in the sense that they are able to represent hierarchically structured knowledge, and can be applied to tasks requiring compositionality (?) [@Kwisthout:2016fx, @Griffiths:2012ii]. In the field of cognitive science, sophisticated model of structured representations using Bayesian networks [@Pearl:1988wz] have been developed [@Tenenbaum:2011hd, @Lucas:2010tu], with applications to concept learning [CITATION], inductive biases in reasoning, in cognitive models of development [@Perfors:2011ex], language [@Goodman:2016go] and social cognition [@Baker:2009kx].
@Kwisthout:2016fx make the point that predictive processing accounts are situated at low levels in the processing hierarchy, usually at the level of sensory inference or at the level describing kinematics. In this paper, we address this issue by starting from a model developed for low-level sensory processing, incorporating basic abilities to predict sensory consequences of movement based on knowledge about the kinematics, and extending this to account for interactions between thought processes, operationalized as mental simulations of sensorimotor experience, and sensory processing.

> TODO: CITE [@Battaglia:2013fm]

To our knowledge, few studies have addressed the idea that higher level cognitive processes are grounded in the circuits performing sensory inference. We claim that if cognition is grounded in sensory inference, and that this is to be experimentally studied by analyzing interactions between thought and perception, then it is vital that this be carried within the constraints of a theoretical (computational) framework. A suitable formalism for this is the framework of probabilistic graphical models, and within this framework, interactions between thought and perception can be modelled as effects of higher level priors and hyperpriors on inference at a computational level. Investigating inference at the level of algorithmic implementations might provide further insight.

@@

We investigate one specific type of head movement, a rotation of the head about the yaw axis, and discuss this in the context of the probabilistic computations that are performed when the goal is to:

1) infer motion caused by external events
2) infer the consequences of one's own actions.

@@

## Previous work
Whereas early models of vestibular sensory processing were inspired by control theory [@Borah:1988fd, @Merfeld:1999cg], several recent studies have described vestibular sensory processing as dynamic Bayesian inference [@Laurens:2006be, @PaulRMacNeilage:2008bv, @Karmali:2012cv]. @Selva:2012gsa
provide a recent review of the relationship between optimal observer models and models that employ Kalman filtering, and show that these are largely equivalent. Both @Karmali:2012cv and @Laurens:2006be described vestibular sensory inference in terms of particle filtering, which is a particular algorithm that can be used to perform inference in state-space models. [@Doucet:2000bh, @Doucet:2009us]. Using their model, @Laurens:2006be were able to simulate perceptual responses to centrifugation and off-vertical axis rotations. A well-known phenomenon of vestibular perception is velocity storage, which is defined as the slower decay of perceptual and oculomotor responses, relative to the SCC afferent responses [@Raphan:1979hn], __[Robinson 1977]__. This is generally seen as evidence for the existence of an internal model in the processing of vestibular signals. @Karmali:2012cv focussed on one-dimensional angular velocity during yaw rotations; their model was able to explain the phenomenon of velocity storage as a consequence of particle filtering, with the time constant of velocity storage depending on the variability of the sensory afferent signals.

## Generative model for sensory inference
We proceed by providing a brief introduction to the concept of state-space models, and particle filtering, which is a particular algorithm that can be used to perform inference in state-space models. While Kalman filters provide the optimal solution when the restrictions of linear Gaussian state-space models are met, they are not easily extensible to non-linear models, which is relevant when considering multi-dimensional motion. Particle filtering can be applied to arbitrary nonlinear models. State-space models can be seen as special cases of more general Bayesian hierarchical models [@Bishop:2006ui; @Murphy:2012ua]. Following the same approach as @Penny:2013iv, we first reframe the generative model used for sensory inference as a dynamic probabilistic graphical model.
 <!-- from a statistical perspective, rather an engineering perspective -->
A state-space model is special case of a dynamic latent variable model, which is suitable for modelling the time-varying behaviour of a system in terms of latent, unobservable processes and noisy observations. As such, a state-space model can be viewed as a statistical model of the problem at hand.

For example, consider trying to estimate the velocity and position during a leftward head turn. The afferent SCC signals are considered to be proportional to the angular velocity of the head with a frequency band of approximately 0.1 – 10 Hz, which constitutes the range of natural head movements [@Grabherr:2008kk; @Carriot:2014fv]. The measurements are therefore observations of the true, unobservable angular velocity; however, they are contaminted by Gaussian noise, which is determined by characteristics of the sensory organs. Figure \@ref(fig:head-turn-data) shows a sequence of noisy measurements during a two-second turn of the head. The acceleration generating the movement is sinusoidal, given for each time step $t$ by $Asin(2\pi t/T)$, where $A$ is the amplitude of the movement, $T = 1/f$ is the duration of the movement, and $f$ is the frequency. This acceleration results in a peak velocity $\omega_{peak} = AT/\pi$ and a final angular position $\theta_{final} = AT^2/2\pi$.  Due to the inherent noisiness of the measurements, and the dynamic nature of the stimulus, the measurements need to be interpreted according to the context in which they occur.

@@

<!-- \autoref{fig:graphical-model-inference} -->
Figure \@ref(fig:graphical-model-inference) shows a representation of a state-space model as a graphical model, unrolled over multiple time steps. Starting at time 0, the state evolves according to the process model, with the measurements at each time step being generated according to the measurement model. The 2-dimensiontal latent state variable is unshaded, to indicate that it cannot be observed, wheres the observed sensory measurements are shaded. Whereas previous models were discussed in terms of the particular filtering algorithm used, the graphical notation highlights the fact that a state-space model can be viewed as a dynamic latent factor model [@Blei:2014cp]. This opens up the possibility of extending the graphical model to include higher-level variables. 
Given a generative model, we can apply various sampling algorithms, either in order to run a Monte Carlo simulation by sampling from the prior without incorporating any observed variables, or to perform inference, by conditioning on observed variables. After a brief discussion particle filtering, an inference algorithm often used to infer the values of the latent state variables, we will introduce our extension to the model that will allow a prediction of the consequences of active movements.

@@

$$ \omega_{t} = \omega_{t-1} + \alpha_{t-1} \Delta t $$
$$ \alpha_t = \alpha_{t-1} $$

@@

# TODO
How do we explain higher-level anticipation/expectations in such a model? This form of anticipation is not the same as the 1-step ahead predictions made during inference, but rather implement higher level prior beliefs about the direction of movement, for example.

An obvious way of including this is the mechanism that allows one to predict the sensory consequences of self-initiated movement via an efference copy.

@@

__A further issue is that the sensory inference model cannot incorporate any top-down expectations or biases. For example, imagine seeing that someone is going to push you. You can use this information, derived from the visual modality, in order prepare yourself in order to compensate for this. However, you do not want to act too early, but rather wait until you are actually moving.__

@@

We implement the control in the following manner: during an active head movement, the model generates a sequence of control signals $u_{1:T}$. For example, during a leftward head turn, using the sinusoidal acceleration profile shown in Figure \@ref(fig:head-turn-data), the acceleration is given by $Asin(2\pi t/T)$. Assuming the model has full knowledge about the onset of movement, and the type of movement (sinusoidal head turn), the model requires information about the frequency, $f = 1/T$, amplitude, $A$, and direction, which is given by the sign of the amplitude. These are

<!-- Equation XXX [refer to acceleration] -->

@@

Show a table demonstrating the difference between:

- prediction: 1-step ahead during inference
- expectation: higher level priors

@@

- ability to select type of motion
- parameters of the motion

@@

# Various sources of interference

Interference is rather surprising; if the goal is to perform simulation without incorporating the data, then we should run a Monte Carlo simulation. Why should this interfere with sensory inference perform by a particle filter? We propose that the answer is in the shared generative model.

Here, we identify various sources of interference (2-way) between thought (higher-level cognition) and lower-level sensory processing. This interference can be either due to low-level influences in sensory inference, due to priors on higher-level parameters, choice of process model (false expectations of certain type of motion). This is shown by model simulations.

@@

## Summary
Interference may the result of either:

1) mis-specified proposal distribution: inappropriate priors, or wrong type of motion or
2) biased decision-making

We should be able to investigate which by applying cognitive process models that explain response times and choices.

@@

## Model specifications

## 3-d state space model

Figure \@ref(fig:3d-statespace) show a 3-d state space model

```{r 3d-statespace, echo=FALSE, fig.cap="3-d state space model.", message=FALSE, warning=FALSE}

load("simulations/model-fits/m1-3d-statespace.Rda")

df_w <- with(m1_3d_statespace$summ_smc, {
    data_frame(t = 1:length(w$f$mean),
               mean = w$f$mean,
               lower = w$f$quant$`0.025`,
               upper = w$f$quant$`0.975`)
})
df_w$param <- "Velocity"
df_w$true <- m1_3d_statespace$model$data()$w_true
df_w$observations <- m1_3d_statespace$model$data()$y

df_p <- with(m1_3d_statespace$summ_smc, {
    data_frame(t = 1:length(p$f$mean),
               mean = p$f$mean,
               lower = p$f$quant$`0.025`,
               upper = p$f$quant$`0.975`)
})
df_p$param <- "Position"
df_p$true <- m1_3d_statespace$model$data()$p_true
df_p$observations <- m1_3d_statespace$model$data()$y

df <- bind_rows(df_p, df_w)

p1 <- df %>%
    filter(param == "Position") %>%
    ggplot(data = ., aes(x = t)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey90") +

    geom_line(aes(y = true), alpha = 0.9,
              linetype = 4, size = 1.2) +

    geom_line(aes(y = mean), size = 1.6) +
    geom_hline(yintercept = 0, linetype = 2, color = "grey70") +
    scale_x_continuous(limits = c(1, 20),
                       breaks = c(0, 5, 10, 15, 20),
                       labels = c("0", "0.5", "1", "1.5", "2")) +
    # xlab("Time (sec)") + 
    ylab("Angular position (deg)") +
    papaja::theme_apa(base_size = 18) +
    theme(axis.title.x=element_blank()) +
    theme(panel.border = element_blank(), 
          axis.line = element_line(colour="black", size= 0.75, lineend="square"))

p2 <- df %>%
    filter(param == "Velocity") %>%
    ggplot(data = ., aes(x = t)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey90") +

    geom_line(aes(y = true), alpha = 0.9,
              linetype = 4, size = 1.2) +

    geom_line(aes(y = mean), size = 1.6) +
    geom_point(aes(y = observations),
               alpha = 1, fill = "white",
               colour = "white",
               shape = 21, size = 6) +
    geom_point(aes(y = observations), alpha = 1,
               fill = "black",
               colour = "grey40", shape = 21, size = 4) +
    geom_hline(yintercept = 0, linetype = 2, color = "grey70") +
    scale_x_continuous(limits = c(1, 20),
                       breaks = c(0, 5, 10, 15, 20),
                       labels = c("0", "0.5", "1", "1.5", "2")) +
    xlab("Time (sec)") + 
    ylab("Angular Velocity (deg/s)") +
    papaja::theme_apa(base_size = 18) +
    theme(panel.border = element_blank(), 
          axis.line = element_line(colour="black", size = 0.75, lineend="square"))

cowplot::plot_grid(p1, p2, ncol = 1) 
                   # labels = c("Position", "Velocity"),
                   # label_size = 12)

```

## Monte Carlo simulation
Show dynamics when simulating using various proposal distributions implementing various natural head motions:

- turing left
- shaking head quickly
- shaking head slowly (communicative gesture)

## Expectations based on higher level priors
In our model, there is no difference between modelling active self-motion, and an anticpation of passive self-motion due to predictable external forces. This is a testable hypothesis.

## Model mis-specification
### Inappropriate priors

### Wrong type of motion model
For example, expect shaking instead of head turn

```{r active-1, cache=TRUE, fig.cap="Mis-specified proposal distribution.", echo=FALSE, message=FALSE, warning=FALSE}

move <- function(A, t, Time) {
    rnorm(1, A * sin(2 * pi * (t - 1)/Time), 0.1)
}

transfun <- function(x, t, Time, A, sigma, N, dt = 0.1, selfmotion) {
    mu <- matrix(rep(0, 2*N), nrow = N)

    if (selfmotion == "active") {
        mu[, 1] = x[, t-1, 1] + dt * x[, t-1, 2] + dt^2/2 * move(A, t, Time)
        mu[, 2] = x[, t-1, 2] + dt * move(A, t, Time)
    } else if (selfmotion == "passive") {
            mu[, 1] = x[, t-1, 1] +  dt * x[, t-1, 2]
            mu[, 2] = x[, t-1, 2]
        }
        else {
            mu[, 1] = 0
            mu[, 2] = 0
        }
    # TODO: more elegant way of doing this?
    t(apply(mu, 1, function(x) {rmvnorm(1, mean = x, sigma = sigma)}))
}

params <- list(sdx = c(0.8, 0.8) * eye(2), sdy = 2.0,
                x_init = c(0, 0),
                sdx_init = 0.5*eye(2),
                A = 10, N = 1000,
                dt = 0.1, freq = 0.5,
                transfun = transfun)

m1 <- run_particle_filter(motiondata = motiondata, params = params,
                            resample_particles = TRUE, rs_thresh = 0.5,
                            selfmotion = "active")

plot(m1, panel_label = "", print_loglik = FALSE, predict = FALSE, imagine = FALSE)
```

@@

- Nigmatullina
- Wertheim
[@Grabherr:2011ej] showed that patients with bilateral or unilateral vestibular loss show impaired ability to mentally transform images of bodies and body parts.

@@

### Head direction cells
Head direction cells are thought to integrate signals of vestibular origin to maintain a signal of cumulative rotation (reviewed in @Taube:2007fk and are found in many brain areas, including the postsubiculum, retrosplenial cortex, thalamus, lateral mammillary nucleus, dorsal tegmental nucleus, striatum and entorhinal cortex [@Cullen:2014gx]. This may be necessary in order to mentally simulate being in another position, i.e. the brain apparently needs to simulate the n-step dynamics of its motion through space. Why is this? Because vestibular connections make a major contribution to spatial orientation; information about the position in space obtained through the vestibular sensory system is computed by integrating velocity over time.

- mental simulation cannot just re-use the state used in sensory inference, because the velocity is fed to the head direction cells. The brain would get confused
- directional signal is not just a simple integration of velocity signals from the vestibular system
- signals from motor efferecence copy or proprioceptive systems appear to be crucial for upating the preferred firing direction over time

## Testable hypotheses
Show a table of testable hypotheses

### 2017-01-27 notes.Rmd
added Notes, model-fits and new content

# head direction cells

[@Taube:2007fk]

- mental simulation cannot just er-use the state used in sensory inference, because the velocity is fed to the head directino cells. The brain would get confused

- directinal signal is not just a simple integration of velocity signals from the vestibular system
- signals from motor efferecence copy or proprioceptive systems appear to be crucial for upating the preferred firing direction over time
- what about this? you don't know where you are, because you were passively moved. Can you update if I tell you (verbally instruct you)?
- vestibular input is eesntial for establishing the head direction signal? How can we reconcile this with the assumption that re-afference is cancelled, i.e. input due to active motion is cancelled. Maybe it's the control model...

# Corollary discharge
[@Crapse:2008jc]

# Predictive control 
[@Stahl:2011dl]

- movemements can induce sensory input
- sensory input due to own actions is indistinguishable from input due to external forces
- strategy: efference copy -> corollary discharge

it seems intuitively obvious that this mechanism has something to do with mental imagery.
This has been pointed out many times; mental imagery and mental simulation more generelly is somehow related to the capacity to compute corollay discharge, based on the efference copy and a forward model.

> Necessary components:  efference copy and a forward model

Sensory receptors don't care about their input. The brain has to deal with the problem further downstream

>>>>

---
output: 
  html_document: 
    fig_caption: yes
    number_sections: yes
    theme: spacelab
---

# one
that the brain is able to imagine, plan, perform counterfactual simulations because it coopts the mechanism evolved in order to predict the sensory consequences of its actions.

interference arises when he brain, due to task demands, uses the control mechanism in a manner which is not entirely counterfactual

# two
interference is surprising, because a mental simulation should be counterfactual. if it isn't this might be a problem. why should thinking about rotating left have an effect on your ability to sense leftward motion?

we show this type of interactions needs to be investigate pdf in a probabilistic framework

# three
we identify 2 main problem that have not been addressed: in SPT how do people decide which way and for how long to rotate?

and how does imagery interfere?

the first issue speaks to the fact that we have to solve the inverse problem before we can simulate in order to transform our FOR. the is equivalent to solving a planning problem, I.e. choosing the best action in order to reach a goal. this is actually counterfactual inference.

after introducing the graphical model, we show the types of inference, I.e. which nodes are clamped.

then we explain interference as biased perceptual decision making, concentrating on the Nigmatullina effect. how do biases arise? the basic model is not equipped to deal with bias.

# four
but how does interference through CVS on mental spatial transformation work? We hypothesize that CVS is a strange stimulus and needs interpretation by the brain. Passive actual movement puts the brain in an alert mode, because it might have to counteract the motion, I.e. the control circuits are engaged, leading to less capacity for planning.

in any case, we need a closer look at the connection between menral simulation and planning and the ability to predict self initiated movements.

# five
finally, how can we explain the bias in the Nigmatullina experiment? we need to investigate whether it is due to a bias in the starting point, in terms of a traditional method of analyzing such data, i.e. DDM. this is implemented by a prior on the choice alternatives, leading to starting point or altered drift rate.

or is it a model misspecification?

The alternative model by Merfeld would say it's just the starting point.

# six
perform simulations to show a theoretical analysis of what may cause the Nigmatullina effect.

model misspecification might be unlikely, but certainly worth a look

# seven
we tell the story in two parts.

1) describe how we can simulate using a parsimonious model that refuses components of the sensory inference model and components involved in distinguishing active from passive movements.

2) what happens when we simulate. switch off inputs, sensory gating corresponding to not resampling. show tables of correspondence mechanisms. show that interest,S of probabilistic computations the ability to model sensory consequences is a question of creating a proposal distribution for sampling from the prior. show table comparing Monterey carlo simulation with particle filtering. first show tables explaining components of filtering.

# eight
of course we first elucidate how the ability to simulate is related to sensory inference in real time.

vestibular sensory inference is a very good example, because it se ms to be well described as sequential monte carlo

# nine
interference is either:

1) mis-specified proposal distribution
2) biased decision-making

# 10
we focus explicitly know:

- forward simulation, explaining the requirements from a statistical / probabilistic point of view
- interference from thought to inference, and explore how this might occur. we focus on this direction because it is easier to understand than the other direction

show a sequence diagram showing a typical trial in an experiment

# eleven
we implement forward sampling by sampling from the proposal distribution without updating. this corresponds to monte carlo simulation

In order to set up a particular simulation, the brain must create a proposal distribution such that it represents the kinematics of the system to be modelled.
